Vector Representations of Linguistic Communities and Associated Models

### 2021-12-15 paper/paper1.tex
other revisions

The anecdotal evidence suggests that for some communities, proximity
in linguistic patterns do match social proximity and for others it
does not. Unfortunately, it seems hard, if not impossible, to draw a
general conclusion from this mode of analysis. In this section, we propose a
systematic approach with which we can quantify the correlation
between social proximity and linguistic proximity, and measure its
statistical significance.

@@

The problem with this approach is that even if several dimensions of
\(L\) and \(S\) are correlated, they will not coincide in the
\emph{representation}. That is, applying a simple rotation (orthogonal
transformation) on either matrix widely changes the \(||L-S||_F\)
correlation metric. Another way to consider the same problem is that,
applying such a rotation does \emph{not} change the cosine similarities
considered in the previous section.

To make the metric independent of the representation (up to orthogonal
transformations), we compute the \emph{minimum} distance between
\(L_i\) and \(S_i\), for any orthogonal matrix \(\Omega\) applied to
\(L\):

@@

Along these lines, \citet{OConnor2010} uses a Bayesian generative
model to infer communities from variation in text data.  In contrast
to our work, this model treats words as independent events, ignoring
the structure (and variation) in the construction of sequences.  It
does further suggest, however, that community-level variation can be
modeled in an unsupervised way.

There are several recent studies that aim to measure \emph{linguistic distinctiveness}
at the level of speech community \citep{OConnor2010,Zhang2018,Lucy2021}, 
which is an possible interpretation of the community-stratified information gain 
of the CCLM over its unconditioned counterpart \cref{sec:ig}. 
Whereas the metrics in previous work are based on lexical frequency
(and in the case of \citet{Lucy2021}, word sense distributions),
CCLM information gain is capable of capturing 
distinctiveness at multiple levels of linguistic analysis. 
However, further work is needed to
investigate exactly what kinds of variation are captured.

While the focus of the paper is theoretical,
computational models of variation can also support 
robust, equitable language technology.  Previous work has shown that
speaker demographics can improve performance on standard NLP tasks
\citep{Hovy2015,Yang2017}. %\jp{Move this to related work so we can get to the point quicker.}

>>>>

% The anecdotal evidence suggests that for some communities, proximity
% in linguistic patterns do match social proximity and for others it
% does not. Unfortunately, it seems hard, if not impossible, to draw a
% general conclusion from this mode of analysis.

In this section, we propose a
systematic approach with which we can quantify the correlation
between social proximity and linguistic proximity, and measure its
statistical significance.

@@

The problem with the above metric is that even if several dimensions
of \(L\) and \(S\) are correlated, they will not coincide in the
\emph{representation} of embeddings. That is, applying a simple
rotation (orthogonal transformation) on either matrix widely changes
the \(||L-S||_F\) correlation metric.

% Another way to consider the same problem is that,
% applying such a rotation does \emph{not} change the cosine similarities
% considered in the previous section.

To make the metric independent of the representation (up to orthogonal
transformations, which preserve cosine similarities), we compute the
\emph{minimum} distance between \(L_i\) and \(S_i\), for any
orthogonal matrix \(\Omega\) applied to \(L\):

@@

A piece of work with similar goals as ours is that of \citet{OConnor2010}, which uses a Bayesian generative
model to infer communities from variation in text data.  In contrast
to our work, this model treats words as independent events, ignoring
the structure (and variation) in the construction of sequences.  It
does further suggest, however, that community-level variation can be
modeled in an unsupervised way.

There are several recent studies that aim to measure \emph{linguistic distinctiveness}
at the level of speech community \citep{OConnor2010,Zhang2018,Lucy2021}, 
which is an possible interpretation of the community-stratified information gain 
of the CCLM over its unconditioned counterpart (\cref{sec:ig}). 
Whereas the metrics in previous work are based on lexical frequency
(and in the case of \citet{Lucy2021}, word sense distributions),
CCLM information gain is capable of capturing 
distinctiveness at multiple levels of linguistic analysis. 
However, further work is needed to
investigate exactly what kinds of variation are captured.

While the focus of this paper is sociolinguistic aspects,
computational models of variation can also support 
robust, equitable language technology.  Previous work has shown that
speaker demographics can improve performance on standard NLP tasks
\citep{Hovy2015,Yang2017}. %\jp{Move this to related work so we can get to the point quicker.}
